Occultation Light Curves in Reflected Light

### 2020-03-31 tex/ms.tex
more text

%
\section{The Problem}
\label{sec:the-problem}

As in \citet{Luger2019}, we compute the instantaneous visible flux $f$ from
a spherical body whose surface intensity is described by the spherical harmonic
coefficient vector $\mathbf{y}$ as
%
\begin{align}
    \label{eq:sTARRy}
    f = \bvec{s}^\top \bvec{A} \bvec{R}' \bvec{R} \bvec{y}
    \quad,
\end{align}
%
where $\bvec{R}$ and $\bvec{R}'$ are spherical harmonic rotation matrices,
$\bvec{A}$ is a change-of-basis matrix to a basis $\tilde{\bvec{g}}$,
and $\bvec{s}^\top$ is the \emph{solution vector}, the integral of each
of the basis functions in $\tilde{\bvec{g}}$ over
the projected visible disk of the body (which may be partially occulted by
another spherical body).
For more information on these terms, the reader is referred to
\citet{Luger2019}.

\begin{figure}[t!]
    \begin{centering}
        \includegraphics[width=0.35\linewidth]{figures/illum.pdf}
        \oscaption{illum}{%
            Illumination geometry.
            \label{fig:illum}
        }
    \end{centering}
\end{figure}

The derivations in \citet{Luger2019} assume that the coefficient vector
$\mathbf{y}$ describes the \emph{emissivity} of the body, which (in the
absence of limb darkening) is assumed to be Lambertian, i.e., all points on the
surface emit equally in all directions.
Here, we derive the solution to the occultation problem for Lambertian
reflectance, in which case the vector $\mathbf{y}$ describes the
surface \emph{reflectivity} (or \emph{Bond albedo}). Specifically, we assume
the body is illuminated by a point-like source and the radiance at any point
is proportional to the cosine of the angle between the incident light
and the surface normal. Points for which this angle is greater than
$\frac{\pi}{2}$ are unilluminated and therefore have radiance of zero;
see Figure~\ref{fig:illum}.

% case 1: 0
% case 2: sTr
% case 3: sTe * I + ~sTr
% case 4: sTe * I
% case 5: sTr - sT0 * I
% case 6: (sTe + sT0) * I + ~sTr
% case 7: (sTe - sT0) * I
% case 8: sT0 * I

@@

\section{Cases 1--4}
\label{sec:cases-1-4}
TODO.

\section{Cases 5--8}
\label{sec:cases-5-8}

>>>>

%
\section{The Problem}
\label{sec:the-problem}
%
As in \citet{Luger2019}, we compute the instantaneous visible flux $f$ from
a spherical body whose surface intensity is described by the spherical harmonic
coefficient vector $\mathbf{y}$ as
%
\begin{align}
    \label{eq:sTARRy}
    f = \bvec{s}^\top \bvec{A} \bvec{R}' \bvec{R} \bvec{y}
    \quad,
\end{align}
%
where $\bvec{R}$ and $\bvec{R}'$ are spherical harmonic rotation matrices
\citep[\S3.3 of][]{Luger2019},
$\bvec{A}$ is the change-of-basis matrix from spherical harmonics
to a basis $\tilde{\bvec{g}}$ \citep[Equation B13 in][]{Luger2019},
and $\bvec{s}^\top$ is the \emph{solution vector}, the integral of each
of the basis functions in $\tilde{\bvec{g}}$ over
the projected visible disk of the body (which may be partially occulted by
another spherical body).
%
The basis $\tilde{\bvec{g}}$ is called the \emph{Green's basis}; its name
stems from the fact that its components have a structure that makes integration
by Green's theorem convenient. Its components were defined in
Equation~11 of \citet{Luger2019b}:
%
\begin{proof}{bg}
    \tilde{g}_{l,m} &=
    \begin{dcases}
        %
        \frac{\mu+2}{2}x^\frac{\mu}{2} y^\frac{\nu}{2}
         & \qquad \mu, \nu \, \mathrm{even}
        \\[1em]
        %
        z(x, y)
         & \qquad \mu = \nu = 1
        \\[1em]
        %
        3x^{l-2}yz(x, y)
         & \qquad \nu \, \mathrm{odd}, \,
        \mu = 1, \,
        \frac{\mu + \nu}{2} \, \mathrm{even}
        \\[1em]
        %
        z(x, y)
        \bigg(
        -x^{l-3} + x^{l-1} + 4x^{l-3}y^2
        \bigg)
         & \qquad \nu \, \mathrm{odd}, \,
        \mu = 1, \,
        \, \mathrm{odd}
        \\[1em]
        %
        z(x, y)
        \bigg(
        \frac{\mu-3}{2} x^\frac{\mu-5}{2} y^\frac{\nu-1}{2}
        \ - \
        \frac{\mu-3}{2} x^\frac{\mu-5}{2} y^\frac{\nu+3}{2}
        \\
        \qquad\qquad \ - \
        \frac{\mu+3}{2} x^\frac{\mu-1}{2} y^\frac{\nu-1}{2}
        \bigg)
         & \qquad \mathrm{otherwise}
        \quad,
    \end{dcases}
    \label{eq:bg}
\end{proof}
%
where $x$ and $y$ are the Cartesian components on the surface
of the projected
disk of the body, with $z(x, y) = \sqrt{1 - x^2 - y^2}$, and
%
\begin{align}
    \mu & = l - m
    \nonumber     \\
    \nu & = l + m
    \quad.
\end{align}
%
The vector $\tilde{\bvec{g}}$ is structured such that
the component at index $n$ is $\tilde{g}_{l,m}$ with
\begin{align}
    \label{eq:lm}
    l & = \floor*{\sqrt{n}} \nonumber \\
    m & = n - l^2 - l
    \quad.
\end{align}
\xxx{The rotation matrices do the following...}
For more information on all of these terms, the reader is referred to
\citet{Luger2019}.

\begin{figure}[t!]
    \begin{centering}
        \includegraphics[width=0.35\linewidth]{figures/illum.pdf}
        \oscaption{illum}{%
            Illumination geometry for the unocculted case.
            A point source at Cartesian coordinates
            $(0, y_s, z_s)$ illuminates a uniform,
            Lambertian, spherical body of radius unity. The observed
            radiance falls off as the cosine of the angle $\phi$ between the
            source and the surface normal. Beyond the terminator
            (defined by $\phi = \frac{\pi}{2}$), the radiance is zero. In
            projection, the terminator is a segment of an ellipse of semi-major
            axis unity and semi-minor axis
            $b = -\nicefrac{z_s}{\sqrt{y_s^2 + z_s^2}}$.
            \label{fig:illum}
        }
    \end{centering}
\end{figure}

The derivations in \citet{Luger2019} assume that the coefficient vector
$\mathbf{y}$ describes the \emph{emissivity} of the body, which (in the
absence of limb darkening) is assumed to be Lambertian, i.e., all points on the
surface emit equally in all directions.
Here, we derive the solution to the occultation problem for Lambertian
reflectance, in which case the vector $\mathbf{y}$ describes the
surface \emph{reflectivity} (or \emph{Bond albedo}). Specifically, we assume
the body is illuminated by a point-like source and the radiance at any point
is proportional to the cosine of the angle $\phi$ between the incident light
and the surface normal. Points for which $\phi \ge \frac{\pi}{2}$ are
unilluminated and therefore have radiance of zero; see Figure~\ref{fig:illum}.

In \citet{Luger2019}, the crux of the problem was to compute the terms in the
vector $\bvec{s}^\top$, which are surface integrals over
the region of the occulted body that is visible to
the observer (Figure 2 in that paper). Here, these integrals are significantly
more difficult, because they involve not only the intersection of the
limbs of the two bodies, but also intersections with the elliptical
day/night terminator of the occulted body. Figure~\ref{fig:cases} shows the
eight families of cases involving an occultation of an illuminated body by
a spherical occultor. Each of these cases correspond to distinct ways in
which the limb of the occultor can intersect with the limb of the occulted
body and its terminator; together, these cases encompass all possible
occultation configurations, for any illumination angle, occultor size, and
occultor position. In the next three sections, we present the solution for the
$\bvec{s}^\top$, and hence for the total visible flux, for each of these cases.

% case 0: sTr
% case 1: 0
% case 2: sTr
% case 3: sTe * I + ~sTr
% case 4: sTe * I
% case 5: sTr - sT0 * I
% case 6: (sTe + sT0) * I + ~sTr
% case 7: (sTe - sT0) * I
% case 8: sT0 * I

@@

\section{The Solution: Case 0}
\label{sec:cases-0}
%
Before we tackle configurations involving occultations, we must address the
simpler problem of computing the total visible flux from an unocculted
body in reflected light (see Figure~\ref{fig:illum}). This problem was
originally solved by \citet{Haggard2018} and subsequently adapted to the
\starry formalism in \citet{Luger2019b}, but for completeness we present
the detailed solution below.

Our task is to compute the solution vector for this
configuration, which we will denote $\bvec{s}^\top_0$.  When no occultor
is present (as in this case), it is
more convenient to perform the integration in the \emph{polynomial basis}
$\tilde{\bvec{p}}$ \citep[Equation 7 in][]{Luger2019}:
%
\begin{align}
    \tilde{p}_{l,m}(x, y) & =
    \begin{dcases}
        x^\frac{\mu}{2} y^\frac{\nu}{2}
         & \qquad \mu, \nu \, \mathrm{even}
        \\[1em]
        x^\frac{\mu-1}{2} y^\frac{\nu-1}{2} z(x, y)
         & \qquad \mathrm{otherwise} \quad,
    \end{dcases}
\end{align}
%
where \xxx{we unroll p as before...}
The solution vector is then given by
%
\begin{align}
    \label{eq:sTref}
    \bvec{s}^\top_0 & = \bvec{r}^\top \bvec{A_1} \bvec{R} \bvec{A}^{-1}
    \quad,
\end{align}
%
where $\bvec{r}^\top$ is the solution vector in the polynomial basis,
$\bvec{A_1}$ is the change-of-basis matrix between the polynomial basis
and the spherical harmonic basis, $\bvec{R}$ is a spherical harmonic rotation
matrix, and $\bvec{A}$ is the change-of basis matrix between the spherical
harmonic basis and the Green's basis.
\xxx{In words, the equation above reads...}

\xxx{...} In this frame, the body sits at the origin and is illuminated by a point
source in the
$y-z$ plane at coordinates $(0, y_s, z_s)$. It is straightforward to show
that the illumination at a point $(x, y)$ on the projected disk is
given by
%
\begin{proof}{illum}
    \Psi(b; x, y) & =
    \begin{dcases}
        b_cy - bz(x, y) &
        \quad\quad\quad\quad\quad\quad\quad\quad\quad\quad
        y \geq b \sqrt{1 - x^2}
        \\
        0               &
        \quad\quad\quad\quad\quad\quad\quad\quad\quad\quad
        \mathrm{otherwise}
    \end{dcases}
    \quad ,
    \label{eq:illum}
\end{proof}
%
where $b_c = \sqrt{1 - b^2}$.
%
\xxx{The solution vector may then be computed as...}
%
\begin{align}
    \label{eq:rT}
    r_{n(\mu,\nu)} & =
    \int_{-1}^{1}
    \int_{b\sqrt{1 - x^2}}^{\sqrt{1 - x^2}}
    \tilde{p}_{n(\mu,\nu)}(x, y)
    \Psi(b; x, y)
    \ \dd y \ \dd x
    \quad.
\end{align}
%
Equation~(\ref{eq:rT}) may be solved analytically in terms of purely
trigonometric and algebraic functions of $b$:
%
\begin{proof}{rT}
    \label{eq:rTsoln}
    r_{n(\mu,\nu)} & =
    \begin{cases}
        \frac{b_c\left(1 - b^{\frac{\nu + 4}{2}}\right)}{2}
        J_{\frac{\mu}{2}, \frac{\nu + 2}{2}} -
        b I_\frac{\nu}{2}(b) K_{\frac{\mu}{2}, \frac{\nu}{2}}
        %
         &
        %
        \qquad
        \mu, \nu \ \mathrm{even}
        %
        \\[1em]
        %
        b_c
        I_{\frac{\nu + 1}{2}}(b) K_{\frac{\mu - 1}{2}, \frac{\nu + 1}{2}}
        %
        \\[0.5em]
        \qquad
        %
        - \frac{b}{2} \bigg\{
        \left(1 - b^{\frac{\nu + 1}{2}}\right)
        \left(
        J_{\frac{\mu - 1}{2}, \frac{\nu - 1}{2}} -
        J_{\frac{\mu + 3}{2}, \frac{\nu - 1}{2}}
        \right)
        \\[0.5em]
        \qquad\qquad
        -
        \left(1 - b^{\frac{\nu + 5}{2}}\right)
        J_{\frac{\mu - 1}{2}, \frac{\nu + 3}{2}}
        \bigg\}
        %
         &
        %
        \qquad
        \mathrm{otherwise}
        %
    \end{cases}
\end{proof}
%
where
%
\begin{proof}{rT}
    \label{eq:IJK}
    I_{j}(b) &= \int_b^1 a^j \sqrt{1 - a^2} \dd a
    \nonumber \\
    %
    J_{i,j} &=
    \frac{
        \Gamma\left(\frac{i + 1}{2}\right)
        \Gamma\left(\frac{j + 1}{2}\right)
    }
    {
        \Gamma\left(\frac{i + j + 4}{2}\right)
    }
    %
    \nonumber \\
    %
    K_{i,j} &=
    \frac{
        \Gamma\left(\frac{i + 1}{2}\right)
        \Gamma\left(\frac{j + 4}{2}\right)
    }
    {
        \Gamma\left(\frac{i + j + 5}{2}\right)
    }
    \quad.
\end{proof}
%
Given initial conditions
%
\\[1em]
\begin{minipage}{.33\linewidth}
    \begin{proof}{rT}
        I_{0}(b) &= \frac{\arccos(b) - bb_c}{2}
        %
        \nonumber \\
        %
        I_{1}(b) &= \frac{b_c^3}{3}
        %
        \nonumber
    \end{proof}
\end{minipage}%
\begin{minipage}{.32\linewidth}
    \begin{proof}{rT}
        J_{0,0} &= \pi
        %
        \nonumber \\
        %
        J_{0,1} &= \frac{4}{3}
        %
        \nonumber
    \end{proof}
\end{minipage}%
\begin{minipage}{.33\linewidth}
    \begin{proof}{rT}
        \label{eq:IJK0}
        K_{0,0} &= \frac{4}{3}
        %
        \nonumber \\
        K_{0,1} &= \frac{3\pi}{8}
    \end{proof}
\end{minipage}
\\[1em]
%
we may compute all higher order terms from the recurrence relations
%
\begin{proof}{rT}
    \label{eq:IJKrec}
    I_{j}(b) &= \frac{b^n b_c^3 + (j - 1) I_{j - 2}(b)}{j + 2}
    %
    \nonumber \\
    %
    J_{0,j} &= \left(\frac{j - 1}{j + 2}\right) J_{0,j-2}
    %
    \nonumber \\
    %
    K_{0,j} &= \left(\frac{j + 2}{j + 3}\right) K_{0,j-2}
    %
    \nonumber \\
    %
    J_{i,j} &= \left(\frac{i - 1}{i + j + 2}\right) J_{i-2,j}
    %
    \nonumber \\
    %
    K_{i,j} &= \left(\frac{i - 1}{i + j + 3}\right) K_{i-2,j}
    \quad.
\end{proof}
%

\section{The Solution: Cases 1--4}
\label{sec:cases-1-4}
Cases 1--4 (see Figure~\ref{fig:cases}) involve configurations in which the
occultor does not intersect with
the terminator of the occulted body, and are therefore significantly easier to
solve.

\section{The Solution: Cases 5--8}
\label{sec:cases-5-8}
Cases 5--8 (see Figure~\ref{fig:cases}).
